---
title: "HW2"
format: html
---

The following analysis and visualization aims to answer the question: *How do FEMA National Risk Index scores for counties in California compare to those in other states?* It is completed as part of an assignment for EDS240: Data Visualization & Communication.

**Data:**



```{r}
#| message: false
#| warning: false
# Load necessary libraries
library(tidyverse)
library(here)
library(janitor)
library(ggplot2)
library(ggridges)
library(gghighlight)
library(patchwork)
```

```{r}
#| message: false
#| warning: false
# Read in data
nri_raw <- read_csv(here("data/National_Risk_Index_Counties_807384124455672111.csv")) %>% 
  clean_names()

# Clean data
nri <- nri_raw %>% 
  
  # Keep only 50 US states (no territories)
  filter(state_name %in% state.name) %>% 
  
  # Find mean and median NRI score by state
  group_by(state_name) %>% 
  mutate(state_mean_nri = mean(national_risk_index_score_composite),
         state_median_nri = median(national_risk_index_score_composite)) %>% 
  ungroup() %>% 
  
  # Arrange states by mean state NRI
  mutate(state_name = fct_reorder(.f = state_name, # which group you want to reorder
                                      .x = state_mean_nri),  # order by the mean column we just created
         
  # Create new column with values by median order
           rank = dense_rank(desc(state_mean_nri))) %>% 
  
  # Add column for whether state is California
  mutate(is_ca = case_when(
    state_name_abbreviation == "CA" ~ 1,
    TRUE ~ 0)) %>% 
  
  # Keep only necessary columns
  select(state_name, state_name_abbreviation, state_mean_nri, state_median_nri,
         county_name, county_type, 
         national_risk_index_score_composite, national_risk_index_state_percentile_composite, is_ca, rank) 
  
# Create data frame with just the top 5 states, by median
nri_top5 <- nri %>% 
  filter(rank %in% seq(1, 5, by = 1)) %>% 
  mutate(state_name = reorder(state_name, rank))
```

```{r}
# Save theme colors
ca_red <- "#B72037"
other_states <- "#4175B8"
```


```{r}
mean_plot <- ggplot(data = nri, aes(x = state_mean_nri, y = reorder(state_name, state_mean_nri), color = factor(is_ca))) +
  geom_point() + 
  geom_linerange(aes(xmin = 0, xmax = state_mean_nri)) + 
  labs(x = " ",
       y = " ", 
       title = " ") +
  theme_minimal() +
  scale_color_manual(values = c(other_states, ca_red)) + 
  theme(legend.position = "none",
        axis.text.y = element_text(size = 7),
        plot.title = element_text(face = "bold", hjust = 0.5, size = 11)) 
```

```{r}
top5_boxplot <- ggplot(nri_top5, aes(x = national_risk_index_score_composite, fill = state_name)) +
  geom_boxplot() +
  labs(x = " ",
       y = " ",
       title = "Distributions of States with Highest Means") +
  facet_wrap(~state_name, 
             nrow = 5) +
  scale_fill_manual(values = c(other_states, other_states, ca_red, other_states, other_states)) + 
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
        strip.text.x = element_text(angle = 0, hjust = 0,
                                    size = 10, color = "black"),
        axis.text.y = element_blank()) +
  ylim(-.75, .75)
```

```{r}
mean_plot + top5_boxplot +
  plot_annotation(
  caption = "National Risk Index",
  theme = theme(plot.caption = element_text(hjust = 0.5, size = 12)))

```
```{r}
top5_boxplot_2 <- ggplot(nri_top5, aes(x = national_risk_index_score_composite, fill = state_name)) +
  geom_boxplot() +
  labs(x = " ",
       y = " ",
       title = " ") +
  facet_wrap(~state_name, 
             ncol = 5) +
  scale_fill_manual(values = c(other_states, other_states, ca_red, other_states, other_states)) + 
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
        strip.text.x = element_text(angle = 0, hjust = 0,
                                    size = 5, color = "black"),
        axis.text.y = element_blank()) +
  ylim(-.75, .75) +
  coord_flip()
```



```{r}
mean_plot + inset_element(
  top5_boxplot_2,
  left = 0.5,   # adjust this value to control horizontal position
  bottom = 0.025,
  right = 1,
  top = 0.7    # adjust this value to control height
)
```
## Responses

**1. What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?**

**2. How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?**

**3. Summarize your main finding in no more than two sentences.**

**4. What modifications did you make to this visualization to make it more easily readable?**

**5. Is there anything you wanted to implement, but didn’t know how? If so, please describe.**
