---
title: "EDS240 Homework 2: FEMA NRI Visualization"
author: "Sofia Sarak

format: html
---

The following analysis and visualization aims to answer the question: *How do FEMA National Risk Index scores for counties in California compare to those in other states?* It is completed as part of an assignment for EDS240: Data Visualization & Communication.

## Data and Wrangling

FEMA ([Federal Emergency Management Agency](https://www.fema.gov/about)) has maintained National Risk Index (NRI) data since 2021. The index “provides information for communities most at risk to 18 different natural hazards. It offers a baseline risk measurement for expected annual loss, social vulnerability and community resilience at the Census tract or county level. The data helps to validate, measure and better understand your community’s natural hazard risk.” Risk is defined as the potential for negative impacts resulting from natural hazards. This analysis uses NRI composite scores, which adds together the risk from all 18 hazard types.

```{r}
#| message: false
#| warning: false
# Load necessary libraries
library(tidyverse)
library(here)
library(janitor)
library(ggplot2)
library(ggridges)
library(gghighlight)
library(patchwork)
```

```{r}
#| message: false
#| warning: false
# Read in data
nri_raw <- read_csv(here("data/National_Risk_Index_Counties_807384124455672111.csv")) %>% 
  clean_names()

# Clean data
nri <- nri_raw %>% 
  
  # Keep only 50 US states (no territories)
  filter(state_name %in% state.name) %>% 
  
  # Find median NRI score by state
  group_by(state_name) %>% 
  mutate(state_median_nri = median(national_risk_index_score_composite)) %>% 
  ungroup() %>% 
  
  # Arrange states by median state NRI
  mutate(state_name = fct_reorder(.f = state_name, # which group you want to reorder
                                      .x = state_median_nri),  # order by the median column we just created
         
  # Create new column with values by median order
           rank = dense_rank(desc(state_median_nri))) %>% 
  
  # Add column for whether state is California
  mutate(is_ca = case_when(
    state_name_abbreviation == "CA" ~ 1,
    TRUE ~ 0)) %>% 
  
  # Keep only necessary columns
  select(state_name, state_name_abbreviation, state_median_nri,
         national_risk_index_score_composite, national_risk_index_state_percentile_composite, 
         is_ca, rank) 

  
# Create separate data frame with just the top 10 states, by median
nri_top10 <- nri %>% 
  filter(rank %in% seq(1, 10, by = 1)) %>% 
  mutate(state_name = reorder(state_name, rank))
```

## Creating Visuals

My visualization will have two components: a lollipop chart comparing median NRI of all 50 states and also boxplots of the 5 states with the highest median NRIs.

```{r}
# Create and save theme colors
ca_red <- "#B72037"
other_states <- "#4175B8"
```

Lollipop chart using median NRI:

```{r}
lollipop_medians <- 
  
  # Plot median state NRI, ordering by median and color by the variable `is_ca`, to highlight California
  ggplot(data = nri, aes(x = state_median_nri, y = reorder(state_name, state_median_nri), color = factor(is_ca))) +
  
  # Add lollipop chart components
  geom_point() + 
  geom_linerange(aes(xmin = 0, xmax = state_median_nri)) + 
  
  # Remove axis labels, they will be redundant in combined plot
  labs(x = "National Risk Index",
       y = " ",
       title = "Median NRI of US States") +
  
  # Set/adjust theme and colors
  theme_minimal() +
  scale_color_manual(values = c(other_states, ca_red)) + 
  theme(legend.position = "none",
        axis.title.x = element_text(size = 9),
        axis.text.x = element_text(size = 5),
        plot.title = element_text(face = "bold", hjust = 0.5, size = 11)) 
```

```{r}
top10_boxplot <- 
  
  # Plot just the states with the highest median NRIs, ordering my median
  ggplot(nri_top10, aes(x = national_risk_index_score_composite, y = reorder(state_name, state_median_nri), fill = state_name)) +
  
  # Add boxplot geom and labels (keeping redundant labels blank)
  geom_boxplot() +
  labs(x = "National Risk Index",
       y = " ",
       title = "Distributions of Top 10, by Median") +
  
  # Add custom color scale to highlight California
  scale_fill_manual(values = c(ca_red, other_states, other_states, 
                              other_states, other_states, other_states, 
                              other_states, other_states, other_states, other_states)) + 
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
        axis.title.x = element_text(size = 9)) +
  
  # Switch y axis labels to the right
  scale_y_discrete(position = "right")
  
  top10_boxplot
```

```{r}
top10_boxplot + lollipop_medians +
  plot_layout(axis_titles = "collect_y") +
  plot_annotation(
  title = "Comparison of NRI County Distribution and Median, by US State",
  caption = "Data info",
  theme = theme(plot.caption = element_text(hjust = 0, size = 5),
                plot.title = element_text(face = "bold")))

```

## Responses

**1. What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?**

My variables of interest are:
- state name (categorical)
- NRI (the composite National Risk Index score, which is numeric), by county
- median NRI, by state, which I calculated in my wrangling step

**2. How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?**

Because the question specified *counties* in California, I figured that a distribution of NRI scores was the most appropriate, as it consolidated but retained county-level data, by state. From there, I considered histograms, density plots (using `geom_density_ridges()`), and boxplots, but I found that plotting all 50 states created a visual that I personally thought was too busy. Distributions of NRI scores for all 50 states is an overwhelming amount of information, regardless of which plot type it is presented in. Therefore, I decided to pivot.

I felt that comparing California to 10 states was an adequate number -- it was large enough to see the gradual decrease of medians, but small enough to where it wasn't overwhelming. A big factor using the top 10 states by median was also that California had the highest median score, ensuring its spot on the plot. However, since the assignment asked for a comparison relative to *all* 50 states, I knew I had to incorporate them somehow. I decided to create a lollipop chart to display state medians, because I feel like it is less overwhelming than both boxplots and a bar chart. It also showed the decrease in medians nicely, with California still at the top.

I also opted to use median as my statistic in the lollipop chart because it is represented in the boxplots, as well. It felt important to have that sort of consistency in my two plots.

**3. Summarize your main finding in no more than two sentences.**

California has the highest median NRI out of all 50 US states, with most counties in the 75-100 range.

**4. What modifications did you make to this visualization to make it more easily readable?**

I colored California with its own red color using a Boolean column in the lollipop chart. With the boxplots, I cheated a little and hand-filled each one, since I knew which one exactly was California.

I also switched the axes of my boxplot to be on the right of it (instead of the default left), so my audience can more easily see axis labels for both plots at the same time.

**5. Is there anything you wanted to implement, but didn’t know how? If so, please describe.**

I was hoping to inset the boxplot within the empty space in the lower right corner of the lollipop chart. I made a few attempts using the patchwork package, but wasn't able to figure out the proper dimensions and coordinates to make it look right. For a lot of combinations, even if the plot was positioned correctly, the boxplots themselves weren't appearing because the plot panel was too small for the data to appear. It also would have been difficult and/or time-consuming to adjust axis text and title to fit properly within the smaller box, but still be readable.


todo:
- fix axis text in lollipop chart
- add data info
- final look through from assignment description checklist
- aspect-ratio in code chunk yaml
