---
title: "Homework 3: Visualizing FEMA NRI x ACS Data"
author: "Sofia Sarak"
format: html
editor: visual
---

The following document contains code to answer the question: *How does climate hazard risk exposure vary across racial / ethnic groups in California?* It uses data from [FEMA's National Risk Index (NRI)](https://www.fema.gov/flood-maps/products-tools/national-risk-index) and the [American Census Survey (ACS)](https://www.census.gov/programs-surveys/acs/about.html).

```{r}
#| message: false
# Load necessary packages
library(here)
library(tidyverse)
library(janitor)
```

Access ACS data:
```{r}
#| message: false
# Access ACS data
#....Step 1a: see all available ACS variables + descriptions.....
# acs_vars <- tidycensus::load_variables(year = 2023,
#                                        dataset = "acs1")
# 
# #..............Step 1b: import race & ethnicity data.............
# race_ethnicity <- tidycensus::get_acs(
#   geography = "county",
#   survey = "acs1",
#   # NOTE: you may not end up using all these variables
#   variables = c("B01003_001", "B02001_002", "B02001_003", 
#                 "B02001_004", "B02001_005", "B02001_006",
#                 "B02001_007", "B02001_008", "B03002_012",
#                 "B03002_002"),
#   state = "CA", 
#   year = 2023) |>
#   # join variable descriptions (so we know what's what!)
#   dplyr::left_join(acs_vars, by = dplyr::join_by(variable == name)) 
# 
# #.................Step 2: write ACS data to file.................
# readr::write_csv(race_ethnicity, here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

#..................Step 3: read in your CSV file.................
race_ethnicity <- read_csv(here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

```

Access NRI data:
```{r}
#| warning: false
# Load in NRI data
nri_raw <- read_csv(here("data/National_Risk_Index_Counties_807384124455672111.csv")) %>% 
  clean_names()
```

### Data Wrangling
```{r}
# Select only data from California
nri_ca <- nri_raw %>% 
  filter(state_name == "California")
```

NRI and ACS data have different methods of storing county information, so we have to account for that.
```{r}
# Change county column in ACS to match that of NRI
race_ethnicity_clean <- race_ethnicity %>% 
  
  # Split `NAME` column into `county` and `state`
  separate_wider_delim(
    cols = NAME,
    delim = ',',
    names = c('county', 'state')) %>% 
  
  # Remove the word 'County' from observations, keeping only county name
  mutate(county_name = str_remove(county, "County"))
```

```{r}
# Join NRI data to ACS data, using county column
nri_acs <- nri_ca %>% 
  
```

