---
title: "Homework 3: Visualizing FEMA NRI x ACS Data"
author: "Sofia Sarak"
format: html
editor: visual
---

The following document contains code to answer the question: *How does climate hazard risk exposure vary across racial / ethnic groups in California?* It uses data from [FEMA's National Risk Index (NRI)](https://www.fema.gov/flood-maps/products-tools/national-risk-index) and the [American Census Survey (ACS)](https://www.census.gov/programs-surveys/acs/about.html).

```{r}
#| message: false
#| warning: false
# Load necessary packages
library(here)
library(tidyverse)
library(janitor)
```

Access ACS data:

```{r}
#| message: false
# Access ACS data
#....Step 1a: see all available ACS variables + descriptions.....
# acs_vars <- tidycensus::load_variables(year = 2023,
#                                        dataset = "acs1")
# 
# #..............Step 1b: import race & ethnicity data.............
# race_ethnicity <- tidycensus::get_acs(
#   geography = "county",
#   survey = "acs1",
#   # NOTE: you may not end up using all these variables
#   variables = c("B01003_001", "B02001_002", "B02001_003", 
#                 "B02001_004", "B02001_005", "B02001_006",
#                 "B02001_007", "B02001_008", "B03002_012",
#                 "B03002_002"),
#   state = "CA", 
#   year = 2023) |>
#   # join variable descriptions (so we know what's what!)
#   dplyr::left_join(acs_vars, by = dplyr::join_by(variable == name)) 
# 
# #.................Step 2: write ACS data to file.................
# readr::write_csv(race_ethnicity, here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

#..................Step 3: read in your CSV file.................
race_ethnicity <- read_csv(here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

```

Access NRI data:

```{r}
#| warning: false
#| message: false
# Load in NRI data
nri_raw <- read_csv(here("data/National_Risk_Index_Counties_807384124455672111.csv")) %>% 
  clean_names()
```

### Data Wrangling

```{r}
# Select only data from California
nri_ca <- nri_raw %>% 
  filter(state_name == "California") %>% 
  select(county_name,
         national_risk_index_score_composite,
         national_risk_index_rating_composite)
```

NRI and ACS data have different methods of storing county information, so we have to account for that.

```{r}
# Change county column in ACS to match that of NRI
race_ethnicity_clean <- race_ethnicity %>% 
  
  # Split `NAME` column into `county` and `state`
  separate_wider_delim(
    cols = NAME,
    delim = ',',
    names = c('county', 'state')) %>% 
  
  # Remove the word 'County' from observations, keeping only county name
  mutate(county_name = str_trim(str_remove(county, "County"))) %>% 
  
  # Drop original county column
  select(!county)

# Confirm that both of our data sets now include the same counties
length(unique(nri_ca$county_name)) == length(unique(race_ethnicity_clean$county_name))
```

After reading the metadata, it becomes clear that not all California counties have 1-year race/ethnicity estimates, because ACS only calculates these for counties and places (i.e., towns or cities) with populations of 65,000 or more ([ACS Website](https://www.census.gov/data/developers/data-sets/acs-1year.html#:~:text=The%20ACS%201%2Dyear%20estimates,populations%20of%2065%2C000%20or%20more.)).

The visualizations will therefore contain data only from the 42 California counties with ACS estimates, **not** all 58.

NRI and ACS use different years? 2020 and 2023, respectively

```{r}
# Filer NRI data to only contain the same counties as ACS
# nri_ca_42 <- nri_ca %>% 
#   filter(county_name %in% unique(race_ethnicity_clean$county_name))


# Join NRI data to ACS data, using county column
nri_acs <- nri_ca %>% 
  inner_join(race_ethnicity_clean,
             by = join_by(county_name))
```

### Visualization

```{r}
nri_acs %>% 
  group_by(label) %>% 
  summarize(pop = sum(estimate)) %>% 
  
ggplot(aes(x = pop, y = label)) +
  geom_col()
```
- maybe calculate percent of each group using total pop

Races I need: White, Black or African American, American Indian and Alaska Native, Asian, Native Hawaiian and Other Pacific Islander, Some Other Race, Two or More Races, Hispanic or Latino

```{r}
library(ggbeeswarm)

ggplot(nri_acs, aes(y = label, x = national_risk_index_score_composite)) +
  geom_beeswarm()
```
```{r}
nri_acs_group <- nri_acs %>% 
  group_by(label)

ggplot(nri_acs_group, aes(y = label, x = national_risk_index_score_composite)) +
  geom_beeswarm()
```
```{r}
# second join attempt
nri_acs <- race_ethnicity_clean %>% 
  inner_join(nri_ca, by = join_by(county_name))
```

