---
title: "Homework 3: Visualizing FEMA NRI x ACS Data"
author: "Sofia Sarak"
format: html
editor: visual
---

The following document contains code to answer the question: *How does climate hazard risk exposure vary across racial / ethnic groups in California?* It uses data from [FEMA's National Risk Index (NRI)](https://www.fema.gov/flood-maps/products-tools/national-risk-index) and the [American Census Survey (ACS)](https://www.census.gov/programs-surveys/acs/about.html).

```{r}
#| message: false
#| warning: false
# Load necessary packages
library(here)
library(tidyverse)
library(janitor)
```

Access ACS data:

```{r}
#| message: false
# Access ACS data
#....Step 1a: see all available ACS variables + descriptions.....
# acs_vars <- tidycensus::load_variables(year = 2023,
#                                        dataset = "acs1")
# 
# #..............Step 1b: import race & ethnicity data.............
# race_ethnicity <- tidycensus::get_acs(
#   geography = "county",
#   survey = "acs1",
#   # NOTE: you may not end up using all these variables
#   variables = c("B01003_001", "B02001_002", "B02001_003", 
#                 "B02001_004", "B02001_005", "B02001_006",
#                 "B02001_007", "B02001_008", "B03002_012",
#                 "B03002_002"),
#   state = "CA", 
#   year = 2023) |>
#   # join variable descriptions (so we know what's what!)
#   dplyr::left_join(acs_vars, by = dplyr::join_by(variable == name)) 
# 
# #.................Step 2: write ACS data to file.................
# readr::write_csv(race_ethnicity, here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

#..................Step 3: read in your CSV file.................
race_ethnicity <- read_csv(here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

```

Access NRI data:

```{r}
#| warning: false
#| message: false
# Load in NRI data
nri_raw <- read_csv(here("data/National_Risk_Index_Counties_807384124455672111.csv")) %>% 
  clean_names()
```

### Data Wrangling

```{r}
# Select only data from California
nri_ca <- nri_raw %>% 
  filter(state_name == "California") %>% 
  select(county_name,
         national_risk_index_score_composite,
         national_risk_index_rating_composite)
```

NRI and ACS data have different methods of storing county information, so we have to account for that.

```{r}
# Change county column in ACS to match that of NRI
race_ethnicity_clean <- race_ethnicity %>% 
  
  # Split `NAME` column into `county` and `state`
  separate_wider_delim(
    cols = NAME,
    delim = ',',
    names = c('county', 'state')) %>% 
  
  # Remove the word 'County' from observations, keeping only county name
  mutate(county_name = str_trim(str_remove(county, "County"))) %>% 
  
  # Drop original county column
  select(!county)

# Confirm that both of our data sets now include the same counties
length(unique(nri_ca$county_name)) == length(unique(race_ethnicity_clean$county_name))
```

After reading the metadata, it becomes clear that not all California counties have 1-year race/ethnicity estimates, because ACS only calculates these for counties and places (i.e., towns or cities) with populations of 65,000 or more ([ACS Website](https://www.census.gov/data/developers/data-sets/acs-1year.html#:~:text=The%20ACS%201%2Dyear%20estimates,populations%20of%2065%2C000%20or%20more.)).

The visualizations will therefore contain data only from the 42 California counties with ACS estimates, **not** all 58.

NRI and ACS use different years? 2020 and 2023, respectively

```{r}
# Filer NRI data to only contain the same counties as ACS
# nri_ca_42 <- nri_ca %>% 
#   filter(county_name %in% unique(race_ethnicity_clean$county_name))


# Join NRI data to ACS data, using county column
nri_acs <- nri_ca %>% 
  inner_join(race_ethnicity_clean,
             by = join_by(county_name))
```

Calculate % of county population that each racial/ethnic group make up
```{r}
draft <- nri_acs %>% 
  pivot_wider(names_from = label,
              values_from = estimate) %>% 
  select(county_name, national_risk_index_score_composite, national_risk_index_rating_composite, 
         'Estimate!!Total') %>% 
  rename(estimate_total = 'Estimate!!Total') %>% 
  group_by(county_name) %>% 
  mutate(total_pop = estimate_total)


# Create a data frame that defines total population of each county
total_pops <- nri_acs %>% 
  filter(label == "Estimate!!Total") %>% 
  pivot_wider(names_from = label,
               values_from = estimate) %>% 
  select(-c(moe, concept, national_risk_index_score_composite, national_risk_index_rating_composite)) %>% 
  rename(total_pop = 'Estimate!!Total')

# Join data frame to original nri_acs
nri_acs_total_pop <- nri_acs %>% 
  inner_join(total_pops,
             by = join_by(county_name))

# Now, we can use the new data frame to calculate the percentages
# As a check, we should get that where label is Estimate!!Total, our percetage is 100
nri_acs_total_pop <- nri_acs_total_pop %>% 
  mutate(perc_pop = (estimate/total_pop * 100)) %>% 
  
  # And we can clean up the data frame one more time
  select(county_name,
         national_risk_index_score_composite,
         national_risk_index_rating_composite,
         estimate,
         label, concept, 
         total_pop, perc_pop)
```


```{r}

```


### Visualization

```{r}
nri_acs %>% 
  group_by(label) %>% 
  summarize(pop = sum(estimate)) %>% 
  
ggplot(aes(x = pop, y = label)) +
  geom_col()
```

Races I need: White, Black or African American, American Indian and Alaska Native, Asian, Native Hawaiian and Other Pacific Islander, Some Other Race, Two or More Races, Hispanic or Latino

```{r}
library(ggbeeswarm)

ggplot(nri_acs, aes(y = label, x = national_risk_index_score_composite)) +
  geom_beeswarm()
```

```{r}
ggplot(data = nri_acs_total_pop, aes(x = perc_pop, y = national_risk_index_score_composite)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~label, scales = "free")
```

### Separating Based on Race and Ethnicity

The preliminary plot above is a bit overwhelming. Not only are the labels illegible, but it is difficult to compare identities like "Not Hispanic or Latino". To rectify this, I split the data frame into ethnicity and race categories, and omit rows with total population as their label, altogether. After this separation, we can rejoin the plots together.

```{r}
# Filter for only labels that describe race
nri_acs_race <- nri_acs_total_pop %>% 
  filter(label %in% c("Estimate!!Total:!!Black or African American alone",
                      "Estimate!!Total:!!Asian alone",
                      "Estimate!!Total:!!Some Other Race alone",
                      "Estimate!!Total:!!White alone",
                      "Estimate!!Total:!!American Indian and Alaska Native alone",
                      "Estimate!!Total:!!Native Hawaiian and Other Pacific Islander alone",
                      "Estimate!!Total:!!Two or More Races:"))

# Filter for only labels that describe ethnicity
nri_acs_ethnicity <- nri_acs_total_pop %>% 
  filter(label %in% c("Estimate!!Total:!!Not Hispanic or Latino:",
                      "Estimate!!Total:!!Hispanic or Latino:"))
```

```{r}
# Create clear names from labels
race_labels <- c(
  "Estimate!!Total:!!Black or African American alone" = "Black or African American",
  "Estimate!!Total:!!Asian alone" = "Asian",
  "Estimate!!Total:!!Some Other Race alone" = "Some Other Race",
  "Estimate!!Total:!!White alone" = "White",
  "Estimate!!Total:!!American Indian and Alaska Native alone" = "American Indian and Alaska Native",
  "Estimate!!Total:!!Two or More Races:" = "Two or More Races",
  "Estimate!!Total:!!Native Hawaiian and Other Pacific Islander alone" = "Native Hawaiian and Other Pacific Islander"
)

ethnicity_labels <- c(
  "Estimate!!Total:!!Not Hispanic or Latino:" = "Not Hispanic or Latino",
  "Estimate!!Total:!!Hispanic or Latino:" = "Hispanic or Latino"
)
```


Preliminary plots
```{r}
#| fig-asp: 1.5
race_plot <- ggplot(data = nri_acs_race, aes(x = perc_pop, y = national_risk_index_score_composite)) +
  geom_point(col = "#2E86AB") +
  geom_smooth(method = "lm", se = FALSE,
              col = "#92140C") +
  facet_wrap(~label, scales = "free",
             labeller = as_labeller(race_labels),
             nrow = 4) +
  labs(title = "Race", 
       x = "% of Population",
       y = "County NRI Score") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, color = "#BFACAA", face = "bold",
                              size = 20),
    axis.title = element_text(color = "#BFACAA"),
    axis.text = element_text(color = "#BFACAA"),
    strip.background = element_rect(fill = "#2D3047",
                                    color = "#2D3047"),
    strip.text = element_text(color = "white"),
    plot.background = element_rect(fill = "#FFF8F0"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r}
#| fig-asp: 1.5
ethnicity_plot <- ggplot(data = nri_acs_ethnicity, aes(x = perc_pop, y = national_risk_index_score_composite)) +
  geom_point(col = "#2E86AB") +
  geom_smooth(method = "lm", se = FALSE,
              col = "#92140C") +
  facet_wrap(~label, scales = "free",
             labeller = as_labeller(ethnicity_labels),
             nrow = 2) +
  labs(title = "Ethnicity", 
       x = "% of Population",
       y = "County NRI Score") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, color = "#BFACAA", face = "bold",
                              size = 20),
    axis.title = element_text(color = "#BFACAA"),
    axis.text = element_text(color = "#BFACAA"),
    strip.background = element_rect(fill = "#2D3047",
                                    color = "#2D3047"),
    strip.text = element_text(color = "white"),
    plot.background = element_rect(fill = "#FFF8F0"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
Essentially, counties with more Hispanic or Latino populations show a trend of higher NRI scores.

```{r}
library(patchwork)

race_plot + ethnicity_plot
```

### Response Questions

**1. What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?**

**2. How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?**

**3. Summarize your main finding in no more than two sentences.**

**4. What modifications did you make to this visualization to make it more easily readable?**

**5. Is there anything you wanted to implement, but didn’t know how? If so, please describe.**

to do:
- fix widths of final plot
- fix text size in facet labels
- add overall title and description
- answer qs
- overall organize doc
- alt text, title + subtitle
- update readme


